
# Meet the Penguins {#sec-penguins}

## In-class assignment

```{r, include = FALSE}
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(here)
penguins_raw <- read_csv(here("files", "penguins_raw.csv"))
```


## Session plan and goals

Work in pairs where possible; prioritise running code from a clean R session.

- 0–10 min: Preflight (project structure)
- 10–30 min: Task 1 — Create ONE canonical import script and save a reusable dataset
- 30–45 min: Task 2 — Create an analysis script that reuses the saved dataset
- 45–65 min: Task 3 — Create a Quarto report that reuses the saved dataset
- 65–80 min: Task 4 — Debugging drills (paths, objects, columns, hidden state)
- 80–95 min: Task 5 — Extend analysis and export a summary
- 95–115 min: Responsible AI — warnings, vibe coding, prompting, reflection
- 115–120 min: Wrap-up checklist

Deliverables today:
- scripts/00_import_penguins.R creates data/processed/penguins_clean.rds and .csv
- scripts/01_penguins_analysis.R loads the processed data and runs
- 04_penguin_report.qmd renders using the processed data (or sources 00_import_penguins.R)

Success criteria:
- After Session > Restart R, analysis script and report run end-to-end without setwd() or manual Console objects.
- Summary CSV written to data/processed and verified with file.exists().

Note: Quarto background is adapted from [quarto.qmd](https://github.com/UEABIO/5027A/blob/main/quarto.qmd).

# Quarto background (brief)

Quarto combines narrative text (Markdown), code (R/Python/etc.), and results in one reproducible file (.qmd). Rendering runs code top-to-bottom in a fresh R session, so your file must include every library, import, and object creation it uses.

Create a doc in RStudio: File > New File > Quarto Document (title/author/output can be changed in YAML).

Three parts:
- YAML (between --- at top)
- Markdown text
- Code chunks, e.g.:

````

```{r}`r ''`
# Import our data
data <- read_csv("data.csv")
```

````


Useful chunk options: echo, eval, warning, message, fig-width/height, out-width, fig-align. You can set defaults in YAML (execute: …) and override per-chunk with hash pipe lines:



Paths: Prefer here::here() for robust project-root-relative paths across scripts and .qmd.

Common knit issues: object not found (dependence on Console), file not found (bad paths), duplicate chunk labels, typos/missing punctuation.

Micro-task:
- Do: Add to YAML:
  ```
  format:
    html:
      toc: true
      code-overflow: wrap
  ```
  Check: Re-render and confirm TOC and wrapping.

# Preflight (5–10 min)

Ensure directories exist:

```{r, eval = FALSE}
dir.create(here::here("data","raw"),       recursive = TRUE, showWarnings = FALSE)
dir.create(here::here("data","processed"), recursive = TRUE, showWarnings = FALSE)
dir.create(here::here("scripts"),          recursive = TRUE, showWarnings = FALSE)
dir.create(here::here("results"),          recursive = TRUE, showWarnings = FALSE)

cat("Project root:", here::here(), "\n")
```


Callout:
- R is case-sensitive. Paths must match exactly.
- Predict: What does here::here("data","raw","penguins_raw.csv") resolve to on your machine?

# Task 1 — ONE canonical import script (20 min)

Goal: Keep all import + minimal cleaning in ONE place. Save a reusable dataset to data/processed. All other scripts and .qmd files will load the saved data (or source this script).

1) Create: scripts/00_import_penguins.R
2) Add and run:

```{r, eval = FALSE}
# Title: Canonical import for penguins
# Purpose: Import raw data and save a cleaned dataset for reuse.
# Author: <Your Name> | Date: <Today>

# Packages ----
library(tidyverse)
library(here)
library(readr)
library(janitor)

# Import raw ----
raw_path <- here::here("data","raw","penguins_raw.csv")

# If you don't have the raw file yet, run ONCE to create from palmerpenguins:
# install.packages(c("palmerpenguins","janitor"))
# readr::write_csv(
#   janitor::clean_names(palmerpenguins::penguins_raw),
#   here::here("data","raw","penguins_raw.csv")
# )

stopifnot(file.exists(raw_path))
penguins_raw <- readr::read_csv(raw_path, show_col_types = FALSE)

# Minimal clean ----
penguins_clean <- penguins_raw |>
  janitor::clean_names() |>
  dplyr::mutate(
    culmen_ratio = culmen_length_mm / culmen_depth_mm
  )

# Save reusable dataset ----
processed_csv <- here::here("data","processed","penguins_clean.csv")
processed_rds <- here::here("data","processed","penguins_clean.rds")
readr::write_csv(penguins_clean, processed_csv)
saveRDS(penguins_clean, processed_rds)

```

3) Do → Check
- Do: Source the script or run line-by-line.
- Check: Both penguins_clean.csv and penguins_clean.rds exist in data/processed.

# Task 2 — Analysis script that REUSES the saved data (10–15 min)

Create: scripts/01_penguins_analysis.R

```{r, eval = FALSE}
# Title: Penguins analysis
# Purpose: Explore and summarise morphometrics
# Author: <Your Name> | Date: <Today>
# ----------------------------- ----
# Packages ----
library(tidyverse); library(here); library(readr)

# Data (REUSE the saved dataset) ----
penguins_clean <- readRDS(here::here("data","processed","penguins_clean.rds"))
# Alternative:
# penguins_clean <- readr::read_csv(here::here("data","processed","penguins_clean.csv"))

# Analysis ----
# Add at least one grouped summary

# Plot ----
# Add at least one ggplot
```


Do → Check
- Do: Session > Restart R, then run this script.
- Check: No setwd(); no object not found; summary/plot appear.

# Task 3 — Quarto report that REUSES the saved data (20 min)

Create at the PROJECT ROOT: 04_penguin_report.qmd

```
---
title: "Penguin report"
format:
  html:
    toc: true
    code-overflow: wrap
---
```

```{r, eval = FALSE}
library(tidyverse)
library(here)
library(readr)
```

## Load saved dataset

```{r, eval = FALSE}
# Preferred: load the saved dataset created by scripts/00_import_penguins.R
penguins_clean <- readRDS(here::here("data","processed","penguins_clean.rds"))
# Alternative:
# penguins_clean <- readr::read_csv(here::here("data","processed","penguins_clean.csv"))
```

## Summary

```{r, eval = FALSE}
penguins_summary <- penguins_clean |>
  dplyr::select(species, island, culmen_length_mm, culmen_depth_mm, flipper_length_mm, culmen_ratio) |>
  tidyr::drop_na(culmen_length_mm, culmen_depth_mm, flipper_length_mm) |>
  dplyr::group_by(species, island) |>
  dplyr::summarise(
    mean_flipper_length = mean(flipper_length_mm),
    mean_culmen_ratio   = mean(culmen_ratio),
    .groups = "drop"
  )
penguins_summary
```

## Plot

```{r, eval = FALSE}
library(viridis) # install first if needed
ggplot(penguins_clean,
       aes(culmen_length_mm, culmen_depth_mm, color = species, shape = species)) +
  geom_point(na.rm = TRUE, alpha = 0.8, size = 2.2) +
  scale_color_viridis_d(option = "D", end = 0.92) +
  labs(x = "Culmen length (mm)", y = "Culmen depth (mm)",
       color = "Species", shape = "Species",
       caption = "Source: palmerpenguins (cleaned)") +
  theme_minimal(base_size = 12)
```


Quarto micro‑tasks:
- Change the title and add a figure caption with `#| fig-cap: "Culmen vs depth by species"`.
- Set `#| echo: false` in the plot chunk to hide code.

Do → Check:
- Re-render; confirm TOC, table, plot, and caption display; code is hidden in plot chunk.

# Task 4 — Debugging drills (10–15 min)

Fix each and add a short comment on the bug:

```{r, eval = FALSE}
# A) Path bug
peng <- readr::read_csv("data\\raw\\penguins_raw.csv")

# B) Object name bug
penguins_clean <- janitor::clean_names(penguin_raw)

# C) Hidden-state bug
setwd("/cloud/project/data/raw")
peng <- readr::read_csv("penguins_raw.csv")

# D) Column bug
penguins_clean |>
  dplyr::mutate(culmen_ratio = culmen_length / culmen_depth)
```


# Task 5 — Extend analysis and export (15–20 min)

Finish in scripts/01_penguins_analysis.R:

```{r, eval = FALSE}
# Summaries ----
penguins_culmen_ratio <- penguins_clean |>
  dplyr::group_by(species, island) |>
  dplyr::summarise(
    mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
    mean_culmen_ratio   = mean(culmen_ratio, na.rm = TRUE),
    .groups = "drop"
  )

# Export ----
readr::write_csv(penguins_culmen_ratio,
                 here::here("data","processed","penguins_summary.csv"))

file.exists(here::here("data","processed","penguins_summary.csv"))
```


# Responsible AI: warnings, “vibe coding,” and prompting (15–20 min)

Definition — “Vibe coding”:
- Generating/accepting code because it “looks right,” without understanding or testing.
- Signs: overly verbose pipelines, unused objects, magic numbers, ignoring warnings/errors, piling tweaks until it runs “by luck.”

Risks to watch:
- Hallucinations: invented functions/arguments; wrong package names.
- Out‑of‑date training: advice for old APIs or options.
- Over‑verbosity: unnecessary mutate/select/filter steps that obscure logic.

How to avoid vibe coding:
- Ask for minimal diffs: “Propose the smallest change to fix this and explain why it works.”
- Specify constraints: “R 4.x, tidyverse only, readr IO, no setwd(), use here::here().”
- Verify against docs: `?function`, help pages, package sites.
- Test immediately; keep a clean script; avoid adding packages without need.

Prompting tips (R + tidyverse):
- Provide the smallest reproducible context: error message, a few rows/column names, target outcome.
- State preferences: “Use dplyr verbs; concise, idiomatic code; no data.table.”
- Ask to cite function/argument names you can verify in tidyverse docs.
- Example prompt:
  - “In R 4.3 with tidyverse: This errors `mutate(culmen_ratio = culmen_length / culmen_depth)`. My columns are `culmen_length_mm`, `culmen_depth_mm`. Propose the smallest fix; no new packages; add one-line NA-safe check.”

Good uses:
- Debugging typos/arguments quickly
- Explaining code you don’t recognise
- Generating comments and section headers
- Suggesting small, readable refactors

Reflection (add a short comment in your script or .qmd):
- What did you ask AI? What change did you accept and how did you verify it?
- Did AI propose anything wrong/outdated? How did you detect and correct it?

# Wrap-up checklist (5 min)

- scripts/00_import_penguins.R saves penguins_clean.rds/csv to data/processed.
- scripts/01_penguins_analysis.R runs after Restart R, loading the saved dataset (no setwd()).
- 04_penguin_report.qmd renders and reuses the saved dataset (or sources 00_import_penguins.R).
- data/processed/penguins_summary.csv exists and `file.exists()` is TRUE.
- Plot uses color + shape for accessibility.
- AI usage is documented briefly, including how you verified suggestions.

# Appendix Solutions (collapsed)

Debugging drills — one set of fixes:

```{r, eval = FALSE}

# A) Path bug: prefer here()
peng <- readr::read_csv(here::here("data","raw","penguins_raw.csv"))

# B) Object name bug: correct object name
penguins_clean <- janitor::clean_names(penguins_raw)

# C) Hidden-state bug: remove setwd(), use here()
# setwd("/cloud/project/data/raw")  # REMOVE
peng <- readr::read_csv(here::here("data","raw","penguins_raw.csv"))

# D) Column bug: use correct column names with _mm
penguins_clean |>
  dplyr::mutate(culmen_ratio = culmen_length_mm / culmen_depth_mm)

```


Minimal script that runs after Restart R:

```{r, eval = FALSE}
# Packages ----
library(tidyverse); library(here); library(readr); library(janitor); library(viridis)

# Data ----
penguins_clean <- readRDS(here::here("data","processed","penguins_clean.rds"))
# Alternative:
# penguins_clean <- readr::read_csv(here::here("data","processed","penguins_clean.csv"))

# Analysis ----
penguins_culmen_ratio <- penguins_clean |>
  dplyr::group_by(species, island) |>
  dplyr::summarise(
    mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
    mean_culmen_ratio   = mean(culmen_ratio, na.rm = TRUE),
    .groups = "drop"
  )

# Plot ----
ggplot(penguins_clean,
       aes(culmen_length_mm, culmen_depth_mm, color = species, shape = species)) +
  geom_point(alpha = 0.8, size = 2.2) +
  scale_color_viridis_d(end = 0.92) +
  theme_minimal()

# Export ----
readr::write_csv(penguins_culmen_ratio,
                 here::here("data","processed","penguins_summary.csv"))
```

