# P-value simulation {.unnumbered}

```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(webexercises)
library(here)
```

```{r}
#| include: false

set.seed(20260903)

# Darwin's maize: fifteen paired differences (cross - self, in inches)
darwin <- read_csv(here("files", "darwin.csv")) |>
  pivot_wider(names_from = "type", values_from = "height") |> 
  janitor::clean_names() |> 
  mutate(difference = cross - self)
```

In hour 1 you built a null distribution by flipping signs on cards.
This hour reproduces that procedure in code, then compares it with
the analytic shortcut.


## Coded permutation with prediction {#sec-permutation}

The code below implements the sign-flip randomisation from hour 1.
It flips signs at random on the fifteen paired differences, computes
the mean, and repeats this 1 000 times to build a null distribution.
The two-sided p-value is then computed as the proportion of null
distribution values at least as far from zero as the observed mean.

:::{.task}
::::{.task-header}
**Task: predict before running**
::::

::::{.task-container}

Before you run the code, write down your answers to both questions:

1. What is the observed mean difference? (You can compute this from
   the fifteen values, or recall it from hour 1.)
2. Roughly what p-value do you expect? Use the board from hour 1 as
   your guide.

::::
:::

```{r}
#| label: sign-flip-permutation

differences <- darwin$difference
observed <- mean(differences)

n_shuffles <- 1000

null_dist <- replicate(n_shuffles, {
  signs <- sample(c(-1, 1), length(differences), replace = TRUE)
  mean(differences * signs)
})

# Two-sided p-value
p_two <- mean(abs(null_dist) >= abs(observed))

cat("Observed mean difference:", round(observed, 3), "\n")
cat("Two-sided p-value:      ", round(p_two, 4), "\n")
```

```{r}
#| label: fig-null-distribution
#| echo: false
#| fig-cap: "Null distribution of the mean paired difference under random sign-flipping. The dashed lines mark the observed mean and its mirror."

ggplot(tibble(null = null_dist), aes(x = null)) +
  geom_histogram(bins = 40, colour = "white") +
  geom_vline(xintercept = c(observed, -observed),
             linetype = "dashed", colour = "firebrick") +
  labs(x = "Mean difference under H\u2080",
       y = "Count")
```

:::{.task}
::::{.task-header}
**Task: check your prediction**
::::

::::{.task-container}

Compare the output with your prediction. Where does the observed mean
sit relative to the null distribution in @fig-null-distribution?

`r hide("Answer")`
The observed mean difference is approximately 2.62 inches. Most of
the null distribution sits near zero, because sign-flipping centres
the distribution at zero under H~0~. The observed value falls in the
tail.
`r unhide()`

::::
:::

### Three edits

You will now make three changes to the permutation code above. Each
edit is specified; make them one at a time and re-run after each.

:::{.task}
::::{.task-header}
**Task: increase the number of shuffles**
::::

::::{.task-container}

Change `n_shuffles <- 1000` to `n_shuffles <- 10000`.

Before you run it, predict: will the p-value change substantially, or
will it be similar?

`r hide("Answer")`
The p-value should be similar but more stable. With more shuffles the
estimate of the p-value becomes more precise, but the underlying
proportion it estimates does not change.
`r unhide()`

::::
:::

:::{.task}
::::{.task-header}
**Task: compute the one-sided p-value**
::::

::::{.task-container}

The current code computes a two-sided p-value:

```r
p_two <- mean(abs(null_dist) >= abs(observed))
```

Change this line to compute the one-sided p-value, counting only
values *at least as large* as the observed mean (not the mirror).
Replace the line with:

```r
p_one <- mean(null_dist >= observed)
```

Before you run it, predict: will the one-sided p-value be larger or
smaller than the two-sided value?

`r hide("Answer")`
Smaller. The two-sided p-value counts both tails; the one-sided
p-value counts only one. For a symmetric null distribution, the
one-sided value is approximately half the two-sided value.
`r unhide()`

::::
:::

:::{.task}
::::{.task-header}
**Task: change the statistic to the median**
::::

::::{.task-container}

The code currently uses the mean as the test statistic. Change both
occurrences of `mean` (the observed statistic and the statistic
inside `replicate`) to `median`.

Before you run it, predict: will the p-value change?

`r hide("Answer")`
It may change. The median is a different summary of the data and
responds differently to extreme values. The null distribution changes
shape because it is now built from medians rather than means. The
permutation logic is the same — only the statistic differs.
`r unhide()`

::::
:::


## Short composition: one-sided p-value {#sec-composition}

Reset your code so that it uses 10 000 shuffles, the mean as the
statistic, and the two-sided p-value. The null distribution object
`null_dist` now holds 10 000 sign-flipped means.

:::{.task}
::::{.task-header}
**Task: write the one-sided p-value computation**
::::

::::{.task-container}

Here is the two-sided p-value line:

```{r}
#| eval: false
p_two <- mean(abs(null_dist) >= abs(observed))
```

Using this as a model, write two or three lines of code that:

1. Compute the one-sided p-value (proportion of null distribution
   values greater than or equal to `observed`).
2. Print the result with `cat()`.

Write your code from scratch in the space below. Do not copy and
paste the two-sided line and edit it — write it yourself, using the
model as a reference.

`r hide("Model answer")`
```{r}
#| eval: false
p_one <- mean(null_dist >= observed)
cat("One-sided p-value:", round(p_one, 4), "\n")
```

The key change: remove `abs()` from both sides of the comparison.
You are no longer asking "how far from zero in either direction?" but
"how often does the null distribution exceed the observed value?"
`r unhide()`

::::
:::


## The analytic shortcut {#sec-analytic}

The sign-flip permutation requires no assumptions about the shape of
the data. It builds the null distribution directly from the observed
values. There is a shortcut: if you assume the paired differences
come from a **normal distribution**, then the null distribution has a known
mathematical form — the **t-distribution**. The `t.test()` function
uses this shortcut.

```{r}
#| label: t-test

t.test(darwin$difference, mu = 0, alternative = "greater")
```


:::{.task}
::::{.task-header}
**Task: read the output**
::::

::::{.task-container}

The `t.test()` output contains several numbers. Identify each of the
following in the printed output:

1. The test statistic (labelled `t`).
2. The degrees of freedom (labelled `df`).
3. The p-value.
4. The estimated mean of the differences.

Which of these four have you already computed yourself in this
session, and which are new?

`r hide("Answer")`
The test statistic and degrees of freedom are new — you have not
computed either by hand or in code today. The p-value corresponds to
the one-sided permutation p-value you wrote in @sec-composition
(though the exact value differs because `t.test()` uses the
t-distribution rather than simulation). The estimated mean matches
the `observed` value you computed at the start of @sec-permutation.
`r unhide()`

::::
:::

:::{.task}
::::{.task-header}
**Task: compare p-values**
::::

::::{.task-container}

Compare the p-value from `t.test()` with the one-sided p-value from
your permutation.

1. Are they similar?

2. Why might they differ slightly?

3. Why might we use the t-distribution rather than generating a null
   distribution by simulation every time?

`r hide("Answer")`
1. They should be broadly similar.
2. The permutation p-value is an estimate based on a finite number of
   shuffles; the t-test p-value is exact under its assumptions.
   Additionally, the t-test assumes normally distributed differences,
   which may not hold exactly.
3. The parametric assumption (normality of the differences) allows the
   null distribution to be computed mathematically rather than
   simulated. That assumption gives us a shortcut, rather than having to brute force a distribution through simulation. When the
   assumption holds, the two approaches agree. When it does not, the
   permutation can give a more trustworthy answer.
`r unhide()`

::::
:::


## Alpha and error, by simulation {#sec-alpha}

What happens when the null hypothesis is actually true? The simulation
below runs two thousand studies in which there is genuinely no effect
(the true mean difference is zero). For each study, it draws fifteen
differences from a normal distribution centred at zero, runs a
t-test, and records whether the p-value falls below 0.05.

```{r}
#| label: alpha-simulation

library(tidyverse)

set.seed(123)

n_studies <- 2000

simulated <- tibble(
  study = 1:n_studies,
  p_value = replicate(n_studies, {
    fake_differences <- rnorm(
      15,
      mean = 0,
      sd = sd(darwin$difference)
    )

    t.test(fake_differences, mu = 0)$p.value
  })
) %>%
  mutate(
    significant = p_value < 0.05
  )

# Summary
cat(
  "Studies where p < 0.05:",
  sum(simulated$significant),
  "out of",
  n_studies,
  "\n"
)

cat(
  "Proportion:",
  round(mean(simulated$significant), 4),
  "\n"
)

```


```{r}
# Visualise the p-value distribution
ggplot(simulated, aes(x = p_value)) +
  geom_histogram(
    binwidth = 0.025,
    boundary = 0,
    colour = "white"
  ) +
  geom_vline(
    xintercept = 0.05,
    linetype = "dashed",
    linewidth = 1
  ) +
  annotate(
    "text",
    x = 0.05,
    y = Inf,
    label = "α = 0.05",
    hjust = -0.1,
    vjust = 1.5
  ) +
  labs(
    title = "False positives under the null hypothesis",
    subtitle = paste0(
      n_studies,
      " simulated studies; n = 15 per study"
    ),
    x = "p-value",
    y = "Number of simulated studies"
  ) +
  theme_minimal(base_size = 13)
```




:::{.task}
::::{.task-header}
**Task: interpret the false-alarm rate**
::::

::::{.task-container}

1. In every one of these two thousand studies, the null hypothesis was
   true. How many studies nevertheless produced a p-value below 0.05?
   
2. What name do we give to this threshold?

3. What name do we give to the error of rejecting a true null
   hypothesis?

`r hide("Answer")`
1. Approximately 100 (5% of 2 000), though the exact count will vary
   because the simulation is random.
   
2. **Alpha** ($\alpha$): a threshold chosen in advance, below which
   the p-value is declared significant. It is a property of the
   procedure, not of the data.
   
3. **Type I error**: rejecting the null hypothesis when it is in fact
   true. The long-run rate of Type I errors equals alpha.

A **Type II error** is the reverse: failing to reject the null
hypothesis when it is in fact false. This is developed further in
later sessions.
`r unhide()`

::::
:::

::: {.callout-note}
## Forward link
Anything that inflates the effective sample size — such as treating
non-independent observations as independent — inflates the Type I
error rate. This is the subject of Session 8.
:::


## Change the threshold {#sec-threshold}

:::{.task}
::::{.task-header}
**Task: change alpha from 0.05 to 0.01**
::::

::::{.task-container}

In the alpha simulation above, the threshold is set at 0.05. Change
it to 0.01.

Before you run it, predict: roughly how many of the two thousand
studies will now produce a p-value below 0.01?

`r hide("Answer")`
Approximately 20 (1% of 2 000). The false-positive rate tracks the
threshold: if you set alpha at 0.01, about 1% of studies in which the
null is true will clear it. This confirms that alpha is a property of
the procedure — you choose how often you are willing to be wrong.
`r unhide()`

::::
:::


## Significance is not importance {#sec-significance}

A p-value depends on sample size. With enough data, a trivially small
effect can produce a tiny p-value. Biological importance does not
depend on sample size — a difference of 0.01 mm in plant height is
biologically negligible regardless of how precisely you estimate it.

Consider two experiments on the same species:

- **Experiment A**: mean difference = 0.05 mm, SE = 0.01, n = 10 000,
  p < 0.001.
- **Experiment B**: mean difference = 3.2 mm, SE = 1.8, n = 12,
  p = 0.09.

:::{.task}
::::{.task-header}
**Task: which result matters?**
::::

::::{.task-container}

Before reading on, write down your answer: which experiment has
produced the more biologically important result, A or B? Give your
reason in one sentence.

`r hide("Answer")`
Experiment B. The effect in A (0.05 mm) is trivially small; it
reaches statistical significance only because the enormous sample
size makes the standard error tiny. The effect in B (3.2 mm) is
large enough to be biologically meaningful, but the small sample
leaves too much uncertainty to clear the 0.05 threshold. Statistical
significance tells you that an effect is unlikely to be zero. It does
not tell you that the effect matters.
`r unhide()`

::::
:::


## Cumulative spine and exit ticket {#sec-exit}

Session 1 established that a statistic is random. Session 2 showed
how to quantify uncertainty from a single sample. This session showed
how to judge whether a specific claim is supported. Session 4 asks
how this generalises when the comparison involves more than two
conditions.

:::{.task}
::::{.task-header}
**Task: exit question**
::::

::::{.task-container}

Write your answer on paper before you leave. This question is
deliberately unanswerable from today's material alone — we will
pick this up next time.

> You tested whether two pollination treatments differ in maize
> height and obtained a coefficient of 1.2 with a standard error
> of 0.5. A third pollination treatment now enters the study. Can
> you still use a single difference of means, and if not, what do
> you need?

::::
:::

## Reading 

- ["Hypothesis Testing and the Null Hypothesis"](statisticsbyjim.com/hypothesis-testing/hypothesis-testing/). 


- ["Understanding P-values"](statisticsbyjim.com/hypothesis-testing/interpreting-p-values/).

- ["Statistical Significance"](statisticsbyjim.com/hypothesis-testing/statistical-significance/).

