# Hypotheses - Pre-Reading {.unnumbered}

```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(webexercises)
library(here)
```

Complete this material before your workshop session. 

## Video: hypotheses as population claims {#sec-video}

::: {.callout-note}
## Watch before class

{{< video https://www.youtube.com/embed/0oc49DyA3hU?si=M0ckNvFNjXlw_S7x >}}

After watching, you should be able to write a null and an alternative
hypothesis as statements about a population.

:::

The video introduces the idea that a statistical hypothesis is a claim
about a population, not about a sample. It covers:

- The **null hypothesis**, written H~0~, and the **alternative
  hypothesis**, written H~1~.
- Why the null is a specific, testable claim (usually of no effect or
  no difference) while the alternative states what you would conclude
  if the null were rejected.
- What it means for an alternative to be **directional** (one-sided)
  or **non-directional** (two-sided).

The video does not cover alpha, significance thresholds, or p-values.
Those appear in the workshop.


## Reading: hypotheses and the measurement {#sec-reading}

### Hypotheses are claims about populations

In the last session you estimated a population quantity from a single sample
and quantified your uncertainty with a confidence interval. That
interval told you *where the population value plausibly lies*. It did
not tell you whether a specific claim about the population is
supported by the data. That is the job of a hypothesis test.

A **statistical hypothesis** is a precise claim about a population
parameter. You always work with two hypotheses as a pair.

The **null hypothesis** (H~0~) states that there is no effect, no
difference, or no relationship in the population. It is the claim you
will try to reject. The null must be specific enough to generate
predictions: if H~0~ is true, what data would you expect?

The **alternative hypothesis** (H~1~) states that the population
parameter differs from the null value. It is what you conclude if the
data are sufficiently unlikely under H~0~.

A hypothesis is always about the population. A statement about what
happened in your sample is a description, not a hypothesis.

Is the following a valid null hypothesis? 

"The mean of my sample is 5.2." `r torf(FALSE)`

`r hide("Explanation")`
No. A hypothesis is a claim about a population parameter, not a
description of a sample quantity. The sample mean is a known number;
there is nothing to test.
`r unhide()`

### Directionality

The alternative hypothesis can be **non-directional** (two-sided) or
**directional** (one-sided).

A two-sided alternative states that the population parameter differs
from the null value, without specifying the direction. For a
population mean $\mu$:

$$H_0: \mu = 0 \qquad H_1: \mu \neq 0$$

A one-sided alternative specifies the direction of the difference:

$$H_0: \mu = 0 \qquad H_1: \mu > 0$$

The choice between one-sided and two-sided is made before looking at
the data. It follows from the research question, not from the results.

```{r}
#| echo: false
opts_direction <- sample(c(
  "One-sided, because the question specifies a direction",
  answer = "Two-sided, because the question asks whether the treatment has an effect without specifying a direction",
  "It depends on the results"
))
```

A researcher asks: "Does this drug affect blood pressure?" Which
alternative is appropriate?

`r longmcq(opts_direction)`

`r hide("Explanation")`
The question asks whether the drug *affects* blood pressure, not
whether it raises or lowers it. That is a two-sided question. If the
question were "Does this drug *lower* blood pressure?", the
alternative would be one-sided.
`r unhide()`

### The measurement: Darwin's maize data

You already know this dataset from last time. Darwin grew fifteen
pairs of maize plants, one cross-pollinated and one self-pollinated
in each pair, and measured their heights. For each pair, the quantity
of interest is the **paired difference**: the height of the
cross-pollinated plant minus the height of the self-pollinated plant.

The question is whether cross-pollination increases height. Stated as
hypotheses about the population mean of the paired differences,
$\mu_d$:

$$H_0: \mu_d = 0 \qquad H_1: \mu_d > 0$$

H~0~ claims that cross-pollination makes no difference to height in
the population. H~1~ claims that cross-pollinated plants are taller
on average.

Is this alternative one-sided or two-sided?
`r mcq(c(answer = "One-sided", "Two-sided"))`

`r hide("Explanation")`
One-sided. Darwin's question was specifically whether cross-pollination
*increases* height, not merely whether it changes height. The
alternative specifies a direction: $\mu_d > 0$.
`r unhide()`

Notice that the hypotheses say nothing about the fifteen plants Darwin
actually grew. They are claims about what would happen across all
possible pairs of plants, of which Darwin observed fifteen.

### What comes next

In the workshop you will generate data under the null hypothesis and
use it to judge whether the observed difference is surprising. The
formal name for that judgment is a **p-value**. Before class, complete
the commitment task and self-test below.


## Commitment task {#sec-commitment}

::: {.callout-warning}
## Bring this to class

Write down, in one or two sentences, what you currently think
"statistically significant" means. Do not look it up. There is no
mark for this and no correct answer is expected — the point is to
have something concrete to revise in the workshop.

Bring what you wrote to the session.
:::


## Self-test {#sec-selftest}

::: {.webex-check .webex-box}

**1.** You have a 95% confidence interval of [3.8 - 4.6] for the
difference in height. A colleague claims the true difference is zero.
Does your interval answer whether they are wrong, and if not, what
would?

```{r}
#| echo: false
opts_exit2 <- sample(c(
  "Yes — zero is excluded from the 95% confidence interval, so the data provide evidence against a zero effect.",
  answer = "The interval excludes zero, which at the 5% level is consistent with rejecting the null hypothesis of a zero effect.",
  "No — a confidence interval cannot tell us anything about zero."
))
```

`r longmcq(opts_exit2)`

**2.** A null hypothesis is a statement about
`r mcq(c("a sample", answer = "a population", "a test statistic"))`.

**3.** A researcher asks whether a new fertiliser changes crop yield
compared to the current fertiliser. The appropriate alternative
hypothesis is `r mcq(c("one-sided", answer = "two-sided"))`.

`r hide("Explanation")`
The question asks whether yield *changes*, not whether it increases or
decreases. A non-directional question calls for a two-sided
alternative.
`r unhide()`

**4.** In Darwin's maize experiment, the paired difference is defined
as the height of the `r mcq(c(answer = "cross-pollinated", "self-pollinated"))`
plant minus the height of the
`r mcq(c("cross-pollinated", answer = "self-pollinated"))` plant.

**5.** The choice between a one-sided and two-sided alternative should
be made `r mcq(c(answer = "before looking at the data", "after looking at the data", "after computing the p-value"))`.

`r hide("Explanation")`
The choice between a one-sided and two-sided test should be made before looking at the data, because choosing the direction after seeing the result can bias the inference and inflate the false-positive rate. The direction should instead be determined by the research question and scientific rationale in advance.
`r unhide()`

:::
