# Random sampling {.unnumbered}

```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(webexercises)
library(here)
```


This page summarises what we covered in the first hour of Session 1. Use it for revision; the full activities were worked through in class.

**Progression:** population → sample → statistic → repeat → sampling distribution.

## The unit of observation {#sec-unit-obs}

We began with two scenarios that required identifying the independent unit of observation. The key distinction is between **biological replicates** (independent units that tell you about the population) and **technical replicates** (repeated measurements of the same unit that tell you about measurement precision). The sample size is the number of independent units, not the total number of measurements.

::: {.callout-note}
## The rule
Two observations are independent if knowing the value of one tells you **nothing about the value of the other**, beyond what you already knew from the population. When multiple measurements come from the same biological unit, the unit is the observation.
:::

::: {.webex-check .webex-box}

**Check your understanding.** A researcher measures the wing length of 30 individual butterflies, measuring each butterfly twice (left and right wing). The sample size for wing length is `r fitb("30")`, not `r fitb("60")`.

:::

## The sampling activity {#sec-sampling}

Every student ran the [Human Benchmark](https://humanbenchmark.com/tests/reactiontime) reaction-time test once and noted the average it reported, in milliseconds. That single number was the student's observation, so the **unit of observation** was the student, and the class as a whole was our **population**.

**Round 1: bench groups.** Each bench (5 to 8 students) computed the mean of its members' reaction times and posted it on the class board, together with the number of students in the group.

**Round 2: random samples.** Every student then entered their reaction time and an anonymous ID into the shared class spreadsheet. Each bench drew a random sample of 6 students from the sheet, computed the mean, and posted it on the board as a second row.

::: {.callout-note}
## Drawing the sample in R
With the sheet loaded as `rt_data`, one row per student, a random sample of 6 students and its mean is:
```{r}
#| eval: false
rt_data |>
  slice_sample(n = 6) |>
  summarise(mean_rt = mean(rt_ms))
```
:::

The population mean was then computed from every student's value on the sheet and revealed. No group's mean, in either row, was exactly equal to it. The sample mean varies from sample to sample because it depends on which students happened to be included. This is why a statistic is random.

::: {.callout-warning}
## The mean of the group means is not the population mean
The benches were different sizes, so averaging the bench means gives each student in a small bench more weight than each student in a large one. The population mean must be computed from the individual values, or from the bench means weighted by bench size.
:::

Three features of the board matched the predictions from the pre-class reading, though with only a handful of points per row the third is hard to judge in the room; the computation session shows it properly:

- Both rows were **centred** near the population mean, the random row more convincingly so.
- The means were **narrower** in spread than the individual reaction times, though wider than most students expected. With only 6 students per sample, averaging removes less of the variability than it would with 20.
- The individual reaction times were right-skewed (a few slow students, no very fast ones); the means were less so.

## Imprecision versus bias {#sec-bias}

A sample mean can be wrong in two independent ways.

**Imprecision** means the sample mean is far from the population mean by chance. A larger sample reduces imprecision — the sampling distribution narrows as $n$ increases.

**Bias** means the sampling method systematically pushes the estimate in one direction. A larger sample does not fix bias; it only makes a wrong answer more precise.

The two rows on the board differed in sampling method, not in sample size. A bench is not a random sample of the class: people sit with friends, friends tend to be similar in age and habits, and a whole bench may be on phones while another is on laptops. A bench mean therefore reflects who sits together as well as chance. The random draws in round 2 removed that influence, which is why the second row is the one to trust to be centred on the population mean.

::: {.webex-check .webex-box}

**Check your understanding.**

A bench of eight students who all play fast-paced video games computes a mean of 231 ms. Adding more students from the same friendship group will move the mean towards the population mean. `r torf(FALSE)`

```{r}
#| echo: false
opts_bias <- sample(c(
  "imprecision, because the bench mean will sometimes be too high and sometimes too low",
  answer = "bias, because the way the bench was formed systematically favours faster students",
  "neither, because the population is the same regardless of how you sample"
))
```

The difference between that bench's mean and the population mean is mainly an example of `r longmcq(opts_bias)`

A random sample of 6 students gives a mean of 290 ms, well above the population mean. Drawing 20 students instead of 6 would be expected to give a mean closer to the population mean. `r torf(TRUE)`

:::

## Connecting the two halves {#sec-connection}

Both the unit-of-observation exercise and the sampling activity address the same idea. Each student was one unit of observation; the five trials behind each student's reported average were technical replicates, and we used their mean, not the trials themselves. Each group of students gave you one statistic. A different group would have given a different statistic. The board was a collection of those statistics — the sampling distribution made visible.

## Exit ticket {#sec-exit}

The question posed at the end of the session:

> You have one sample mean of 272 ms from 6 students. What else do you need before you can say how far this is likely to be from the population mean?


