# Linear models in R {.unnumbered}

```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(webexercises)
library(here)
```

```{r}
#| include: false

set.seed(20260826)

darwin <- read_csv(here("files", "darwin.csv"))

# Compute paired differences (Cross - Self within each pot)
darwin_diff <- darwin |>
  pivot_wider(names_from = type, values_from = height) |>
  mutate(difference = Cross - Self)
```

In hour 1 you ranked scenarios by strength of evidence and arrived at
the t-statistic as the ratio of estimate to standard error. Here you
fit the linear model in R and read its coefficient table.

The data are the same paired differences you have worked with since
Session 2: the height of the cross-pollinated plant minus the height
of the self-pollinated plant within each pot.

## The paired differences {#sec-data}

:::{.task}
::::{.task-header}
**Task: inspect the paired differences**
::::

::::{.task-container}

Run the code below to see the fifteen paired differences. These are
the same values you used in Sessions 2 and 3.

::::
:::

```{r}
#| label: show-differences

darwin_diff |>
  select(pair, Cross, Self, difference)
```

## Fit the intercept-only model {#sec-fit}

:::{.task}
::::{.task-header}
**Task: predict the intercept**
::::

::::{.task-container}

Before running any code, write down your prediction. You have computed
the mean of these differences before. What value do you expect the
intercept to be?

::::
:::

```{r}
#| label: lm-paired

lsmodel1 <- lm(difference ~ 1, data = darwin_diff)
```

The formula `difference ~ 1` means "fit an intercept only." The
intercept is the mean of the paired differences. The coefficient
table tests whether that mean differs from zero — the same question
the permutation test in Session 3 answered.

## Read the coefficient table {#sec-summary}

:::{.task}
::::{.task-header}
**Task: read the summary output**
::::

::::{.task-container}

Run the code below. Compare the intercept with your prediction, then
answer:

1. What does the estimate represent?
2. What does the t-value represent, and how does it relate to the
   ranking exercise from hour 1?
3. Is there evidence that the mean difference is not zero?

`r hide("Answer")`
1. The estimate is the mean of the fifteen paired differences — the
   average advantage of cross-pollination over self-pollination
   within a pot.
2. The t-value is the estimate divided by its standard error. A
   larger absolute t-value means the estimate is further from zero
   in standard-error units, just as in the ranking exercise.
3. Check the p-value. If it is below 0.05, there is evidence that
   the mean difference is not zero, consistent with the permutation
   test from Session 3.
`r unhide()`

::::
:::

::: {.panel-tabset}

## Base R

```{r}
#| label: summary-base

summary(lsmodel1)
```

## Broom

```{r}
#| label: summary-broom

broom::tidy(lsmodel1, conf.int = TRUE)
```

:::

## Confidence interval from the model {#sec-ci}

:::{.task}
::::{.task-header}
**Task: obtain and interpret the confidence interval**
::::

::::{.task-container}

Run the code below. The output is a 95% confidence interval for the
mean paired difference.

1. Does this interval include zero?
2. Compare it with the confidence interval you computed in Session 2.
   Are the values the same (or very close)?

`r hide("Answer")`
1. If the interval does not include zero, this is consistent with the
   p-value being below 0.05.
2. The values should match or be very close. The linear model and the
   manual computation from Session 2 use the same formula: estimate
   ± critical value × SE.
`r unhide()`

::::
:::

::: {.panel-tabset}

## Base R

```{r}
#| label: ci-base

confint(lsmodel1)
```

## Broom

```{r}
#| label: ci-broom

broom::tidy(lsmodel1, conf.int = TRUE)
```

:::

::: {.callout-note}
## A simple write-up
From the linear model on paired differences: "Cross-pollinated maize
plants were taller on average than self-pollinated plants, with a
mean paired difference of 2.62 inches (95% CI: 0.004, 5.22;
t(14) = 2.15, p = 0.05)."

Note: the degrees of freedom are 14 (fifteen pairs minus one), not
28. Match the values to your output.
:::

## Compare with the one-sample t-test {#sec-compare-ttest}

:::{.task}
::::{.task-header}
**Task: confirm the equivalence**
::::

::::{.task-container}

Run the code below. Compare the p-value from `t.test()` with the
p-value from `summary(lsmodel1)`. They should be identical.

This confirms that the intercept-only linear model on paired
differences is the same test as a one-sample t-test. The linear model
is the more general framework.

::::
:::

```{r}
#| label: ttest-paired

t.test(darwin_diff$difference, mu = 0)
```

## Composition task: fit the unpaired model {#sec-composition}

:::{.task}
::::{.task-header}
**Task: write an `lm()` call for the unpaired comparison**
::::

::::{.task-container}

You have just fitted a model that respects the paired design:

```r
lsmodel1 <- lm(difference ~ 1, data = darwin_diff)
```

Now write an `lm()` call that ignores the pairing entirely: model
height as a function of pollination type, using the original
`darwin` data frame. Assign the result to `lsmodel2` and run
`summary()` on it.

`r hide("Hint")`
The response is `height`, the predictor is `type`, and the data are
`darwin`. The formula takes the form `response ~ predictor`.
`r unhide()`

`r hide("Model answer")`
```r
lsmodel2 <- lm(height ~ type, data = darwin)
summary(lsmodel2)
```
`r unhide()`

::::
:::

:::{.task}
::::{.task-header}
**Task: compare the two models**
::::

::::{.task-container}

Look at the coefficient table for `lsmodel2`.

What does the `typeSelf` coefficient estimate? Is this the same
quantity as the intercept of `lsmodel1`?


`r hide("Answer")`

The `typeSelf` coefficient estimates the difference in mean height
   between self-pollinated and cross-pollinated plants — the same
   quantity as the intercept of `lsmodel1`, but with the opposite
   sign (Self minus Cross rather than Cross minus Self). The absolute
   value is the same.

`r unhide()`

::::
:::

::: {.callout-note}
## Why pairing matters
Darwin paired his plants by parent because genetic lineage might play a large role in natural variance in height. By computing the difference *within each pair*, the paired
analysis minimises this source of variation and isolates the effect of
pollination type. The unpaired analysis ignores this structure. The lesson: a model should respect the design that
generated the data.
:::

## Exit ticket {#sec-exit}

:::{.task}
::::{.task-header}
**Task: exit question**
::::

::::{.task-container}

Write your answer on paper before you leave. This question is
deliberately unanswerable from today's material alone — we will
pick this up next time.

> You receive a collaborator's spreadsheet to analyse. The "species"
> column contains entries recorded as "Drosophila", "drosophila",
> "D. melanogaster", and "Dros.". You fit
> `lm(wing_length ~ species, data = df)` and get four coefficients
> for species (plus the intercept). Is the model answering the
> question you intended, and what should you do before fitting it?

::::
:::

## Reading

### Library

- The New Statistics with R, 2nd Edition, **Chapter 6** by Andy Hector


### Online

- [Everything is a Linear Model] (danielroelfs.com/posts/everything-is-a-linear-model/), by Daniel Roelfs

- [Common statistical tests are linear models](lindeloev.github.io/tests-as-linear/), by Jonas Kristoffer Lindeløv - *Covers more than we got to today*
