# Building the confidence interval {.unnumbered}

```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(webexercises)
library(here)
```

This page summarises what we covered in the first hour of Session 2.
Use it for revision; the full activities were worked through in class.

**Progression:** SD of the data → SE of the mean → confidence interval.

## Recovering the mechanism from Session 1 {#sec-board-recovery}

We began by returning to the photograph of the Session 1 board. Row 1 showed
every student's own reaction time. Row 3 showed the means of random samples of
6 students drawn from the container of slips, and the spread of the Row 3 dots
was a small picture of the sampling distribution of the mean.

Two numbers were displayed alongside the photograph:

- the SD of the individual reaction times ms,

- the SD of the random-sample means

The first was larger. Averaging cancels out extreme values: a mean of six
students cannot be as extreme as the most extreme single student, so the sample
means cluster more tightly than the individual times. We used Row 3 rather than
the Row 2 bench means because the benches were not random samples and were of
different sizes.

::: {.callout-note}
## The rule
The standard deviation of the sampling distribution of the mean is the
**standard error of the mean** (SE). It is still a standard deviation, but of a
different distribution: the distribution of sample means rather than of
individual values. From a single sample, the SE is estimated from the sample SD:

$$
\text{SE} = \frac{\text{SD}}{\sqrt{n}}
$$
:::



::: {.webex-check .webex-box}

**Check your understanding.** A colleague reports the standard deviation
of blood pressure readings from 100 patients. This number describes
`r mcq(c(answer = "the spread of readings across patients", "the precision of the sample mean", "the width of a 95% confidence interval"))`.

`r hide("Explanation")`

The SD of the readings describes how variable blood pressure is from
one patient to the next. The SE (which is $\text{SD}/\sqrt{100}$)
would describe the precision of the sample mean.

`r unhide()`

:::

## Computing the standard error for Darwin's maize {#sec-se-maize}

We then applied the formula to the dataset introduced in the pre-class
reading: the 15 paired height differences from Darwin's maize
experiment.

You were given the standard deviation of the 15 differences (4.72
inches). At the bench you computed the standard error as:

$$
\text{SE} = \frac{4.72}{\sqrt{15}} \approx 1.22 \text{ inches}
$$

If Darwin repeated the whole experiment many times, each time with 15 new
pairs, the mean differences would be centred on the true mean difference and
would have a standard deviation of about 1.22 inches. @fig-maize-sampling shows
the curve you sketched in class. Darwin's own mean of 2.62 inches is one point
somewhere under this curve; because the true mean is unknown, we cannot say
where.

```{r, echo = FALSE, eval = TRUE}

#| label: fig-maize-sampling
#| fig-cap: "Sampling distribution of the mean difference for samples of 15 pairs. The centre is the unknown true mean; the shaded region is the middle 95%."
#| echo: false
se_maize <- 4.72 / sqrt(15)
curve_df <- tibble(x = seq(-4, 4, length.out = 400) * se_maize) |>
  mutate(density = dnorm(x, sd = se_maize))

ggplot(curve_df, aes(x, density)) +
  geom_area(data = filter(curve_df, abs(x) <= 1.96 * se_maize), fill = "grey85") +
  geom_line() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_x_continuous(breaks = c(-2, -1, 0, 1, 2) * se_maize,
                     labels = c("−2 SE", "−1 SE", "true mean", "+1 SE", "+2 SE")) +
  labs(x = "Mean difference (1 SE = 1.22 inches)", y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

```


::: {.webex-check .webex-box}

**Check your understanding.** A dataset has SD = 8 and $n$ = 64. The SE
is `r fitb(1)`.

`r hide("Explanation")`

$\text{SE} = 8 / \sqrt{64} = 8 / 8 = 1$.

`r unhide()`

:::

## From standard error to confidence interval {#sec-se-to-ci}

We paused to think about what the SE is a ruler for. The point estimate
is the mean of the 15 differences. The SE tells us how far this
estimate is likely to sit from the true mean. The question is: what
range of values for the true mean is consistent with the estimate we
obtained?

The sampling distribution of the mean is approximately normal, and 95%
of a normal distribution lies within about two standard deviations of
its centre. Since the SE is the standard deviation of the sampling
distribution, we get:

- 95% of sample means fall within about 2 × SE of the true mean.
- Equivalently: the true mean falls within about 2 × SE of about 95% of
  sample means.

The second statement is what the confidence interval expresses.

## Constructing the confidence interval {#sec-ci-construction}

::: {.callout-note}
## The formula
A 95% confidence interval for the mean is:

$$
\bar{x} \pm 1.96 \times \text{SE}
$$

where $\bar{x}$ is the sample mean and SE is the standard error. The
multiplier 1.96 is the exact value from the normal distribution that
puts 95% of the distribution within $\pm 1.96$ SDs of the centre; the
value 2 is close enough for a quick check. For small samples we use a
slightly larger multiplier from the t-distribution, introduced in
Session 3.
:::

Substituting the Darwin numbers on the board:

$$
95\% \text{ CI} = 2.62 \pm 1.96 \times 1.22 \approx [0.2, 5.0] \text{ inches}
$$

Every quantity in this interval was built from what you already had.
The sample mean (2.62) is the average of the 15 differences. The SE
(1.22) is $\text{SD}/\sqrt{15}$. The multiplier (1.96) comes from the
normal distribution.

::: {.webex-check .webex-box}

**Check your understanding.** A study reports a mean of 12.0, an SD of
6.0, and $n$ = 36. The SE is `r fitb(1)` and the approximate 95% CI
runs from `r fitb(10, tol = .1)` to `r fitb(14, tol = .1)`.

`r hide("Explanation")`

$\text{SE} = 6.0/\sqrt{36} = 1.0$. The 95% CI is $12.0 \pm 1.96 \times
1.0$, which is approximately 10.0 to 14.0.

`r unhide()`

:::

## Consolidation: the first eight pairs {#sec-consolidation}

To fix the mechanism, we repeated the construction on a subset of
Darwin's data: the first eight pairs only. You were given the mean and
SD of these eight differences and, at the bench, computed:

1. the SE (SD divided by $\sqrt{8}$),
2. the 95% CI (mean $\pm$ 1.96 $\times$ SE).

The resulting interval was **wider** than the interval from all 15
pairs. This is not a mistake. It reflects the $\sqrt{n}$ relationship
directly: with fewer observations, the SE is larger, so the interval is
wider. Hour 2 confirms this pattern by simulation.

::: {.webex-check .webex-box}

**Check your understanding.** If you compute a 95% CI from a smaller
subset of the same dataset (same SD, smaller $n$), the interval will
typically be
`r mcq(c(answer = "wider", "narrower", "unchanged"))`.

`r hide("Explanation")`

Decreasing $n$ increases the SE ($\text{SE} = \text{SD}/\sqrt{n}$), so
the interval gets wider. This is the same $\sqrt{n}$ relationship you
will confirm by simulation in hour 2.

`r unhide()`

:::

## What the confidence interval does not claim {#sec-ci-warnings}

We closed hour 1 by flagging three plausible-sounding interpretations
of "95% confidence", all of which are wrong. The correct interpretation,
and why these three fail, is the topic of hour 2, where the coverage
simulation makes the 95% claim concrete. Keep the answer you wrote for
the pre-class commitment task to hand; you will use it in hour 2.

The three claims to be wary of:

- 95% probability that the true mean lies in this specific interval,
- 95% of the data lie in this interval,
- 95% chance that a future sample mean will fall in this interval.

Each of these is a natural first guess. None is what a frequentist
confidence interval says.

## Predictions for hour 2 {#sec-predictions}

At the end of the hour you wrote predictions on Part F of the worksheet: the
shape and centre of 10,000 sample means, the SE for $n$ = 5, 20, 80 and 320, the
number of misses among 200 intervals, and the effect of a 99% multiplier. Bring
them, and your sketches, to hour 2. Each one is tested there.

## Exit ticket {#sec-exit}

The question posed at the end of the session (end of hour 2):

> You have a 95% confidence interval of [0.2, 5.0] for the mean
> difference in height between cross-fertilised and self-fertilised
> maize. A colleague claims the true difference is zero: that
> cross-fertilisation has no effect. Does your interval answer whether
> they are wrong, and if not, what would you need?
