# Estimation and uncertainty - Pre-Reading  {.unnumbered}

```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(webexercises)
library(here)
```

Complete this material before your workshop session. It carries the
vocabulary you will need in the room.

## Video: from sample to estimate {#sec-video}

::: {.callout-note}
## Watch before class

{{< video https://www.youtube.com/embed/vikkiwjQqfU?si=CD3yn8c-NpWWMTHG >}}

After watching, you should be able to state that the sample mean is a
point estimate of the population mean.
:::

## Reading: estimation and uncertainty {#sec-reading}

### From sample to estimate {#sec-point-estimate}

In Session 1 you built a sampling distribution on the board. Every group
drew a sample from the same population and computed a mean, and the
class saw that each mean was different. The board showed how sample means
vary — and that they cluster around the population mean.

Now turn the reasoning around. In practice you do not know the
population mean. You have one sample and one mean. That single number is
your **point estimate** of the population mean — the best single guess
you can make from the data you have.

A point estimate is not the population value. It is one draw from the
sampling distribution you saw on the board. The question this session
answers is: how far away from the population mean is your estimate
likely to be?

The answer depends on two things: how variable the individual values
are, and how many of them you measured. The next two sections make that
precise.

### Darwin's maize {#sec-darwin}

The dataset we use for the next three sessions comes from an experiment
by Charles Darwin, reported in 1876 and later reanalysed by R. A.
Fisher.

Darwin grew 15 pairs of *Zea mays* (maize) plants. Within each pair,
one plant was **cross-fertilised** (pollen from a different plant) and
one was **self-fertilised** (pollen from the same plant). The two plants
in each pair were grown in the same pot, so any difference in growing
conditions was shared. Darwin then measured the final height of every
plant, in inches.

The biological question is whether cross-fertilisation produces taller
plants. The statistical question — which we will reach in Session 3 — is
whether the observed difference in height is larger than you would
expect from chance alone. For now, the dataset gives us a concrete set
of measurements to work with when we distinguish different ways of
summarising uncertainty.

`r hide("Check your understanding")`

Darwin paired each cross-fertilised plant with a self-fertilised plant
in the same pot. Why?

The pairing controls for differences between pots (soil, light,
watering). Any difference in height within a pair can be attributed to
the fertilisation treatment rather than to the growing environment.

`r unhide()`

### The standard deviation {#sec-sd}

Imagine you compute the difference in height (cross minus self) for each
of Darwin's 15 pairs. Some differences are large and positive, some are
small, and a couple are negative. The **standard deviation** (SD)
measures how spread out those 15 differences are around their mean.

A large SD means the individual values are widely dispersed; a small SD
means they cluster tightly. The SD is a property of the data: it
describes how variable the individuals are. Measuring more individuals
does not make them less variable — a population with variable members
remains variable regardless of how many you sample.

The standard deviation is computed as:

$$
\text{SD} = \sqrt{\frac{1}{n-1}\sum_{i=1}^{n}(x_i - \bar{x})^2}
$$

where $\bar{x}$ is the sample mean and $n$ is the sample size. You do
not need to compute this by hand; `sd()` in R does it. What matters is
the interpretation: the SD tells you about the spread of the individual
values, not about the precision of the mean.

The SD describes the `r mcq(c(answer = "spread of the individual values", "precision of the sample mean", "width of the confidence interval"))`.

`r hide("Explanation")`

The SD measures how far individual observations typically fall from the
mean. It says nothing about the precision of the mean itself — that is
the job of the standard error.

`r unhide()`

### The standard error {#sec-se}

The **standard error** (SE) measures something different: how precisely
your sample mean estimates the population mean. You saw in Session 1
that repeating the sampling process gives a different mean each time.
The standard error is the standard deviation of that sampling
distribution — it quantifies how much your estimate would typically
shift if you ran the study again with a new sample of the same size.

The standard error is computed from the standard deviation:

$$
\text{SE} = \frac{\text{SD}}{\sqrt{n}}
$$

This formula says two things. First, the SE inherits the variability of
the data (through SD). Second, it shrinks with sample size (through
$\sqrt{n}$). Both are intuitive: more variable data produce a less
precise estimate, and more data produce a more precise one.

The critical distinction is this: **the SD describes the data; the SE
describes the estimate.** The SD tells you how spread out the individual
heights are. The SE tells you how much the mean height would change if
you repeated the experiment. They answer different questions, and
confusing them is the single most common error in reading biological
papers.

The SE describes the `r mcq(c("spread of the individual values", answer = "precision of the sample mean", "probability that the null hypothesis is true"))`.

`r hide("Explanation")`

The SE measures how much the sample mean would vary across repeated
studies. It describes the estimate, not the data.

`r unhide()`

### The square-root-of-*n* relationship {#sec-sqrt-n}

The $\sqrt{n}$ in the denominator of the SE formula has a concrete
consequence. Suppose Darwin had measured 60 pairs instead of 15 — four
times as many. The SD would remain roughly the same (the plants are no
less variable), but the SE would be halved, because $\sqrt{60} = 2
\times \sqrt{15}$. Four times the data gives twice the precision, not
four times.

This is the **law of diminishing returns for sample size**: each
additional observation contributes less to precision than the one before
it. Doubling the sample size does not double the precision; it improves
it by a factor of $\sqrt{2} \approx 1.4$.

If you quadruple the sample size, the SE is `r mcq(c("quartered", answer = "halved", "unchanged", "doubled"))`.

`r hide("Explanation")`

$\text{SE} = \text{SD}/\sqrt{n}$. Replacing $n$ with $4n$ gives
$\text{SD}/\sqrt{4n} = \text{SD}/(2\sqrt{n})$, which is half the
original SE.

`r unhide()`

## Commitment task {#sec-commitment}

Before you come to class, write one or two sentences on what you
currently think it means to say that a confidence interval is "95%
confident". Do not look this up: use whatever intuition you already
have from prior encounters with the phrase, from other courses, from
papers you have read, or from general usage. Bring the paper (or the
note on your phone) to class.

There is no right answer at this stage. The point is to have your
current thinking on paper before the workshop, so you can see how it
changes once you have built the interval yourself and seen the
mechanism that gives it the 95%.

## Self-test {#sec-selftest}

::: {.webex-check .webex-box}

**1.** You have one sample mean of 4.2 g from 20 beans. What else do
you need before you can say how far this is likely to be from the
population mean?

You need the `r fitb(c("standard error", "SE", "standard error of the mean"), ignore_case = TRUE)` of the sample mean.

`r hide("Explanation")`

The sample mean alone tells you nothing about how precise it is. You
need the standard error, which requires the standard deviation of the
data and the sample size. This was the exit question from Session 1; the
reading above resolves it.

`r unhide()`

**2.** The standard deviation measures the spread of the
`r fitb(c("data", "individual values", "observations", "individual observations"), ignore_case = TRUE)`.
The standard error measures the precision of the
`r fitb(c("sample mean", "mean", "estimate"), ignore_case = TRUE)`.

**3.** A study reports SD = 8 and $n$ = 16. The standard error is
`r fitb(2)`.

`r hide("Explanation")`

$\text{SE} = \text{SD}/\sqrt{n} = 8/\sqrt{16} = 8/4 = 2$.

`r unhide()`

```{r}
#| echo: false
opts_sd_se <- sample(c(
  answer = "The SD stays roughly the same; the SE decreases.",
  "Both the SD and the SE decrease.",
  "The SD decreases; the SE stays the same.",
  "Both the SD and the SE stay the same."
))
```

**4.** If you increase the sample size from 20 to 80, what happens to
the SD and the SE?

`r longmcq(opts_sd_se)`

`r hide("Explanation")`

The SD describes how variable the individual values are. That does not
change when you measure more of them. The SE shrinks with $\sqrt{n}$,
so quadrupling the sample size halves the SE.

`r unhide()`

:::
