# Null distributions and what a p-value is {.unnumbered}

```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(webexercises)
library(here)
```

This page summarises what we covered in the first hour of Session 3.
Use it for revision; the full activities were worked through in class.

**Progression:** null hypothesis → simulate under H~0~ → null
distribution → observed statistic → p-value.


## Writing hypotheses {#sec-hypotheses}

We began by comparing the hypotheses brought from the pre-class task.
Each bench discussed one new biological scenario (not from the
pre-class reading) and wrote the null and alternative hypotheses as
claims about a population. Benches then stated whether their
alternative was directional.

::: {.callout-note}
## The rule
A hypothesis is a claim about a population parameter, never a
description of a sample. The null hypothesis states no effect or no
difference. The alternative states what you would conclude if the null
were rejected. The choice of directionality follows from the research
question and is fixed before looking at the data.
:::

::: {.webex-check .webex-box}

**Check your understanding.** A colleague writes: "H~0~: the sample
means are equal." Is this a valid null hypothesis? `r torf(FALSE)`

`r hide("Explanation")`
No. A null hypothesis must be a claim about a population parameter,
not about sample quantities. The correct form would state that the
population means are equal.
`r unhide()`

:::


## The shuffle {#sec-shuffle}

We used Darwin's fifteen paired differences to build a null
distribution by hand. If the null hypothesis is true and
cross-pollination has no effect, then each paired difference is
equally likely to be positive or negative — the sign carries no
information. The physical procedure was:

1. Each bench received fifteen cards showing the observed paired
   differences.
2. For each card, a coin flip decided whether to keep the original
   sign or reverse it.
3. The mean of the fifteen sign-flipped differences was computed and
   posted on the board.
4. Each bench repeated this three times.

The class board accumulated these means into a distribution of values
that could arise if the null hypothesis were true. This distribution
is the **null distribution**.

::: {.callout-note}
## The rule
A **null distribution** is the distribution of a test statistic under
the assumption that the null hypothesis is true. Each value on the
board represents one outcome from a world in which cross-pollination
has no effect. The observed mean difference is then compared against
this distribution.
:::

::: {.webex-check .webex-box}

**Check your understanding.** In the sign-flip procedure, what
justifies flipping the sign of a paired difference?

```{r}
#| echo: false
opts_flip <- sample(c(
  "The differences are normally distributed",
  answer = "If the null hypothesis is true, each difference is equally likely to be positive or negative",
  "Flipping signs increases the sample size"
))
```

`r longmcq(opts_flip)`

:::


## The reveal and the count {#sec-reveal}

The observed mean difference was marked on the board. We then counted
how many of the class-generated values were at least as far from zero
as the observed value, in both directions (because a value equally
extreme in the negative direction would be equally surprising under
the null). That count, divided by the total number of shuffles, was
the **p-value**.

Two points were stated explicitly at this moment:

1. The two-tail count follows from the alternative hypothesis written
   before the data were seen. A one-sided alternative would count
   only one direction.
2. The class proportion is an estimate of a quantity defined over all
   possible sign-flip assignments, not just the ones the class
   happened to produce.

::: {.callout-note}
## The rule
The **p-value** is the proportion of values in the null distribution
that are at least as extreme as the observed statistic. It is
computed with respect to the alternative hypothesis stated before the
data were examined.
:::

::: {.webex-check .webex-box}

**Check your understanding.** The class produced 90 shuffled means.
Seven were at least as extreme as the observed difference (in either
direction). The two-sided p-value is `r fitb(7/90, num = TRUE, tol = 0.005)`.

`r hide("Explanation")`
$7 / 90 \approx 0.078$. The p-value is the proportion of null
distribution values at least as extreme as what was observed, counting
both tails.
`r unhide()`

:::


## Four statements {#sec-four-statements}

Each student retrieved the sentence they wrote before class ("what
does statistically significant mean?") and committed to one of four
statements about what a p-value is. Discussion then identified which
statement is correct and why the other three are wrong.

The four statements, in the form used in class:

```{r}
#| echo: false
opts_pval <- sample(c(
  "The probability that the null hypothesis is true",
  answer = "The probability of observing data at least as extreme as the actual data, assuming the null hypothesis is true",
  "The probability that the result was due to chance",
  "The probability that a replication study would produce the same result"
))
```

Which of the following correctly states what a p-value is?

::: {.webex-check .webex-box}

`r longmcq(opts_pval)`

:::

The key asymmetry: a p-value gives the probability of the data given
a hypothesis, not the probability of a hypothesis given the data.
This was linked back to the Bayesian stub from Session 2 — the
second reading of an interval answers a different question, and
computing the probability that a hypothesis is true requires a prior,
which the frequentist framework does not supply.

::: {.callout-note}
## The rule
A p-value is the probability of observing data at least as extreme as
the actual data, assuming the null hypothesis is true. It is not the
probability that the null is true, not the probability that the result
was due to chance, and not the probability that a replication would
produce the same result.
:::


## Link to Session 1 {#sec-link-session1}

We noted that both Session 1 and Session 3 built a distribution of
class-generated values on the board. In Session 1, the spread came
from *which beans were drawn* — sampling variability from a real
population. In Session 3, the spread came from *which signs were
assigned* — the variability of a null distribution constructed under
H~0~. The two distributions answer different questions: Session 1
asked how much a statistic varies across samples; Session 3 asks how
much a statistic varies **when the null hypothesis is true**.

::: {.webex-check .webex-box}

**Check your understanding.** The board distribution in Session 1
represents `r mcq(c(answer = "sampling variability", "the null distribution"))`.
The board distribution in Session 3 represents
`r mcq(c("sampling variability", answer = "the null distribution"))`.

:::


## Exit ticket {#sec-exit}

The question posed at the end of the session:

> You tested whether two groups differ and got a coefficient of 1.2
> with a standard error of 0.5. A third group now enters the study.
> Can you still use a single difference of means, and if not, what do
> you need?
