# Simulating uncertainty {.unnumbered}

```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(webexercises)
library(here)
```

```{r}
#| include: false

set.seed(20260902)

# Known population parameters for the simulation exercises.
# These are NOT the Darwin data — we use a known population so we can
# check whether our intervals contain the true mean.
pop_mean <- 50
pop_sd <- 10
```

In the first hour you built a confidence interval from the ground up:
you saw where the SD came from, computed the SE from it, and used the
SE to construct the interval. This hour, you will interrogate the
mechanism by simulation. You will confirm the $\sqrt{n}$ relationship,
watch what the 95% coverage claim actually looks like across many
intervals, and settle what the interval does and does not mean.

## Simulate the sampling distribution {#sec-sampling-dist}

We begin by confirming something you saw on the board in Session 1: when
you draw repeated samples from a population and compute the mean each
time, the resulting distribution has a characteristic shape.

The code below draws 10,000 samples of size 20 from a population with
mean 50 and standard deviation 10, and plots the distribution of sample
means.

:::{.task}
::::{.task-header}
**Task: predict the shape**
::::

::::{.task-container}

Before you run the code, write down your prediction. What shape will the
histogram of sample means have? Will it be centred on 50, or somewhere
else?

::::
:::

```{r}
#| label: fig-sampling-dist
#| fig-cap: "Sampling distribution of the mean for 10,000 samples of size 20."

n <- 20
n_reps <- 10000

sample_means <- replicate(n_reps, {
  samp <- rnorm(n, mean = pop_mean, sd = pop_sd)
  mean(samp)
})

tibble(sample_mean = sample_means) |>
  ggplot(aes(x = sample_mean)) +
  geom_histogram(bins = 50, colour = "white") +
  geom_vline(xintercept = pop_mean, linetype = "dashed", colour = "red") +
  labs(x = "Sample mean", y = "Count")
```

:::{.task}
::::{.task-header}
**Task: check your prediction**
::::

::::{.task-container}

Was the histogram centred where you expected? What shape does it have?

`r hide("Answer")`

The histogram is centred on 50 (the population mean) and is
approximately bell-shaped — a normal distribution. This holds even
though each individual value comes from a normal population, and it
would also hold for many non-normal populations provided the sample size
is large enough.

This result is the **Central Limit Theorem** (CLT): the sampling
distribution of the mean tends towards a normal distribution as the
sample size increases, regardless of the shape of the individual data.
Because the sampling distribution is approximately normal, we can use
the normal distribution to build the confidence interval formula.

`r unhide()`

::::
:::

## Build a confidence interval {#sec-build-ci}

A 95% confidence interval for the mean is:

$$
\bar{x} \pm 1.96 \times \text{SE}
$$

The code below draws a single sample of size 20 from the same
population and computes its 95% confidence interval.

```{r}
#| label: one-ci

n <- 20
one_sample <- rnorm(n, mean = pop_mean, sd = pop_sd)

sample_mean <- mean(one_sample)
sample_se <- sd(one_sample) / sqrt(n)
lower <- sample_mean - 1.96 * sample_se
upper <- sample_mean + 1.96 * sample_se

c(lower = lower, mean = sample_mean, upper = upper)
```

:::{.task}
::::{.task-header}
**Task: does this interval contain the true mean?**
::::

::::{.task-container}

The population mean is 50. Look at your interval. Does it contain 50?

`r hide("Answer")`

Check whether `lower` $\leq$ 50 $\leq$ `upper`. Most of the time it
will, but not always — about 5% of intervals constructed this way will
miss the true mean. You will see this directly in @sec-coverage.

`r unhide()`

::::
:::

## Standard error and sample size {#sec-se-n}

The SE depends on sample size through the $\sqrt{n}$ relationship. The
code below computes the SE from 1,000 repeated samples of a given size
and reports the average.

```{r}
#| label: se-demo

n <- 20

se_values <- replicate(1000, {
  samp <- rnorm(n, mean = pop_mean, sd = pop_sd)
  sd(samp) / sqrt(n)
})

mean(se_values)
```

:::{.task}
::::{.task-header}
**Task: change the sample size and predict**
::::

::::{.task-container}

You will run this code four times, each time with a different value of
`n`. Before each run, write down your prediction: will the average SE
increase, decrease, or stay the same compared to the previous run?

1. Change `n` to **5**. Predict, then run.
2. Change `n` to **20**. Predict, then run.
3. Change `n` to **80**. Predict, then run.
4. Change `n` to **320**. Predict, then run.

`r hide("Answer")`

| $n$ | Expected SE ($\text{SD}/\sqrt{n}$) |
|---|---|
| 5 | $10/\sqrt{5} \approx 4.5$ |
| 20 | $10/\sqrt{20} \approx 2.2$ |
| 80 | $10/\sqrt{80} \approx 1.1$ |
| 320 | $10/\sqrt{320} \approx 0.56$ |

Each quadrupling of $n$ halves the SE. The simulated values will be
close to these but not exact, because each sample has its own SD.

`r unhide()`

::::
:::

:::{.task}
::::{.task-header}
**Task: state the pattern**
::::

::::{.task-container}

In one sentence, describe how the standard error responds to sample
size.

`r hide("Answer")`

The standard error decreases with the square root of the sample size:
quadrupling $n$ halves the SE.

`r unhide()`

::::
:::

## Interval coverage {#sec-coverage}

If the 95% confidence interval formula works as advertised, then 95% of
intervals constructed this way should contain the true population mean.
The code below simulates 200 confidence intervals and plots them. Each
horizontal line is one interval; the dashed vertical line is the true
population mean.

```{r}
#| label: fig-coverage
#| fig-cap: "200 confidence intervals for the mean. Each line is one interval; red intervals miss the true mean."

n <- 20
n_reps <- 200

sim_results <- replicate(n_reps, {
  samp <- rnorm(n, mean = pop_mean, sd = pop_sd)
  m <- mean(samp)
  se <- sd(samp) / sqrt(n)
  c(mean = m, lower = m - 1.96 * se, upper = m + 1.96 * se)
})

sim_ci <- as_tibble(t(sim_results)) |>
  rename(sample_mean = mean) |>
  mutate(
    rep = row_number(),
    contains_true = lower <= pop_mean & upper >= pop_mean
  )

sim_ci |>
  ggplot(aes(
    x = rep, ymin = lower, ymax = upper,
    colour = contains_true
  )) +
  geom_linerange() +
  geom_hline(yintercept = pop_mean, linetype = "dashed") +
  coord_flip() +
  scale_colour_manual(values = c("FALSE" = "red", "TRUE" = "grey40")) +
  labs(x = "Sample", y = "Value") +
  theme(legend.position = "none")
```

:::{.task}
::::{.task-header}
**Task: count the misses**
::::

::::{.task-container}

How many of the 200 intervals miss the true mean (shown in red)? What
proportion is that? Is it close to what you would expect?

```{r}
#| label: coverage-count

sum(!sim_ci$contains_true)
mean(sim_ci$contains_true)
```

`r hide("Answer")`

Approximately 10 of the 200 intervals (about 5%) should miss the true
mean. The proportion that contain it should be close to 0.95. This is
the repeated-sampling interpretation made visible: the procedure
succeeds about 95% of the time.

`r unhide()`

::::
:::

## What "95% confidence" actually means {#sec-ci-interpretation}

You have just seen the 95% claim made visible. About 95% of the 200
intervals contained the true mean, and about 5% missed it. This is the
coverage guarantee of the procedure.

Take out the answer you wrote for the pre-class commitment task on what
"95% confidence" means. Compare it to the correct interpretation below.

::: {.callout-note}
## The repeated-sampling interpretation
If the study were repeated many times and a 95% confidence interval
computed each time, approximately 95% of those intervals would contain
the true population mean. Any single interval either contains the true
value or it does not; you cannot know which. The 95% is a property of
the procedure that generated the interval, not of the specific interval
you computed.
:::

Three common misinterpretations, each incorrect:

1. **"There is a 95% probability that the true mean lies in this
   specific interval."** In the frequentist framework, the true mean is
   a fixed quantity; it is either inside the interval or not.
   Probability attaches to the procedure, not to the parameter.

2. **"95% of the data lie within this interval."** The confidence
   interval is a statement about the mean, not about individual
   observations. A range containing 95% of the individual values would
   be much wider (roughly mean $\pm 2 \times$ SD), and it has a
   different name: a prediction interval.

3. **"If I repeat the experiment, there is a 95% chance the new sample
   mean will fall in this interval."** The interval is a statement
   about the location of the population mean, not about where future
   sample means will land.

```{r}
#| echo: false
opts_ci <- sample(c(
  answer = "If the study were repeated many times, approximately 95% of the intervals so constructed would contain the true population mean.",
  "There is a 95% probability that the true population mean lies within this specific interval.",
  "95% of the observed data values fall within this interval.",
  "If the study is repeated, there is a 95% chance the new sample mean falls in this interval."
))
```

:::{.task}
::::{.task-header}
**Task: identify the correct interpretation**
::::

::::{.task-container}

Which of the following correctly states what a 95% confidence interval
means?

`r longmcq(opts_ci)`

`r hide("Answer")`

The correct answer is the repeated-sampling interpretation. The other
three describe quantities the confidence interval does not measure.

`r unhide()`

::::
:::

## A different question, briefly {#sec-bayesian-stub}

The confidence interval you built answers: "if I repeated this
procedure many times, how often would my interval contain the true
mean?" It does not directly answer: "given the data I have, what is
the probability that the true mean is greater than zero?"

That second question has a different answer, called a **Bayesian
credible interval**, which requires an additional input: a prior
distribution for the parameter before the data were seen. The two
frameworks answer different questions, and both are legitimate. The
course returns to Bayesian inference in Session 9; for now, notice
only that the question exists.

## Coverage at a different level {#sec-coverage-99}

The 1.96 in the confidence interval formula corresponds to a 95%
confidence level. A different confidence level uses a different
multiplier.

:::{.task}
::::{.task-header}
**Task: modify the simulation for 99% confidence**
::::

::::{.task-container}

Look at the code that produced @fig-coverage. Your goal is to change the
confidence level from 95% to 99%.

1. Find the line in the code that determines the confidence level. Which
   number controls it?
2. The 99% critical value from the normal distribution is 2.576. Replace
   the current multiplier with 2.576.
3. Before you run the code, predict: will more or fewer intervals
   contain the true mean? Will the intervals be wider or narrower?
4. Run the modified code and check your predictions.

`r hide("Hint")`

The multiplier 1.96 appears in the lines that compute `lower` and
`upper`. Replace 1.96 with 2.576 in both lines.

`r unhide()`

`r hide("Answer")`

At the 99% level, approximately 99% of the intervals should contain the
true mean (about 2 misses out of 200, rather than 10). Each interval is
wider, because the multiplier is larger: $2.576 \times \text{SE}$ rather
than $1.96 \times \text{SE}$. The price of greater confidence is a less
precise claim.

`r unhide()`

::::
:::

## Exit ticket

:::{.task}
::::{.task-header}
**Task: exit question**
::::

::::{.task-container}

Write your answer on paper before you leave. This question is
deliberately unanswerable from today's material alone — we will pick
this up next time.

> You have a 95% confidence interval of [0.2, 5.0] for the mean
> difference in height between cross-fertilised and self-fertilised
> maize. A colleague claims the true difference is zero: that
> cross-fertilisation has no effect. Does your interval answer whether
> they are wrong, and if not, what would you need?

::::
:::

## Reading 

- ["Difference Between Standard Deviation and Standard Error"](statisticsbyjim.com/basics/difference-standard-deviation-vs-standard-error/). 


- ["Confidence Intervals: Interpreting, Finding & Formulas!"](statisticsbyjim.com/hypothesis-testing/confidence-interval/).

- ["Central Limit Theorem Explained"](statisticsbyjim.com/basics/central-limit-theorem/).
