# Task Four - Paired t-tests {#sec-weekthree}


```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
library(performance)
source("R/booktem_setup.R")
source("R/my_setup.R")

```

## Objective:

In the previous exercise {@sec-linear}, you used R and the `lm()` function to perform a two-sample t-test. In this assignment, you’ll extend that work by conducting a **paired t-test** using the *Darwin plant height* dataset as an example:

Your task is to: 

1. Import and inspect the data

2. Conduct a *paired t-test* 

3. Explore the relationship between paired observations

4. Visualise and interpret your results

5. Present your analysis in a Quarto report {@sec-reports}

## Learning Outcomes

By completing this assignment, you will:

- Apply the concept of paired data to control for within-pair variation.

- Understand the equivalence between a paired t-test and a linear model with a pair factor.

- Build confidence in constructing reproducible analysis workflows using Quarto.

- Communicate results clearly with integrated text, code, and visuals.

## Task Outline

### Set up your Quarto document

- Create a new .qmd file with a clear title, author, and date.

````md
---
title: "My Report"
author: "Your Name"
format: html
  embed-resources: true
execute:
  echo: false
  warning: false
  message: false

---

````

- Load packages

- Read in the dataset:



````md
```{r}`r ''`
#| include: false

library(tidyverse)
library(here)
library(rstatix)     # simple t-tests with tidy output
library(broom)       # tidy model summaries (optional)
library(sandwich)    # robust (heteroskedasticity-consistent) variance estimators
library(lmtest)      # coeftest/coefci for robust inference
library(performance) # easy model checks

darwin <- read_csv(here("data", "darwin.csv"))
```
````

- Include a brief introduction: What question are we testing?


## Explore and understand the data

Before analysis, ensure you understand what each variable represents.
Check-in:

- How are “Cross” and “Self” plants related within pairs?

- Why might it make sense to treat “pair” as a **factor rather than a numeric variable**?

> Question: Does this design look like independent samples, or are observations linked in some way?

## Carry out a paired *t-test*

Test whether there is a significant difference between pairs in plant height between the two pollination types.

Think about:

- How could you account for pairs in your analysis?

- How could you model this using `lm()`?

> You can add covariates to a linear model by including them on the right-hand side of the formula (after ~ ) separated with + 
> e.g. lm(variable ~ predictor_1 + predictor_2)

:::{.callout-note}

This model analyses height based on type *while controlling for pair*. Including pair will produce coefficients for every pair, but for this assignment you can ignore the individual pair coefficients and **focus on the main effect of type** — this tells you the average difference in height between pairs of Cross and Self plants.

:::


## Examine the correlation between paired measurements

Run a correlation test between the Cross and Self values.

Check-in:

- Is there evidence that pairs are correlated?

- How does this justify the use of a paired rather than independent t-test?

:::{.callout-tip collapse = "true"}
# Correlation with rstatix

```{r, eval = FALSE}
darwin |> 
  pivot_wider(names_from = type, 
              values_from = height) |> 
  rstatix::cor_test(`Self`,`Cross`) 
```

:::

## Visualise your data

Create a plot that includes:

- Raw data (all individual points),

- (*Optionally*) Paired connections (lines between pairs),

- A summary geom (mean ± SE/95% CI or boxplots).

:::{.callout-tip collapse = "true"}
## Jitter plots

If you jitter your points, make sure your lines use the same jitter so they connect correctly (you can share a `position_jitter` object between `geom_line()` and `geom_point()`).

[See here](https://martinasladek.co.uk/content/blog/2022_feb_til-matching-position-jitter-of-points-and-lines-in-ggplot2)

:::

## Summarise your findings

In 4–6 sentences, interpret your analysis:

- Write a short results summary - it should include estimates of difference, uncertainty values, test statistics and p-values.

- How does the estimate of height difference compare to the two-sample t-test in the previous chapter {@sec-linear}?

- What is the strength of correlation between paired observations?

- How do your visuals complement your findings?

:::{.callout-important collapse = "true"}
## Discussion point

- Discuss with your classmates - what effect does adding pair have about our confidence on the effect of inbreeding?

- Should we exclude variables that are a part of our study design if they make us less confident about the results?

:::



## Rendering your Quarto doc

After completing your analysis, render your `.qmd` file to HTML to produce a polished report. Check the following:

- Code runs top to bottom – all chunks execute without errors.

- Figures and tables appear correctly – plots are visible, captions make sense.

- Formatting is clear – headers, text, and code chunks are readable.

- Interpretation is included – your narrative explains results, not just code output.
    
    - Include a figure
    - Include a short results section (with a model write-up and correlation test)

:::{.callout-warning collapse = "true"}
## My document won't render

If the document doesn’t render, check that all packages are loaded and all objects are created in previous chunks.

Check out {@sec-reports} for more help.
:::
