# Task Six - Interaction models {#sec-weeksix}


```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
library(performance)
source("R/booktem_setup.R")
source("R/my_setup.R")

```

## Objective:

In the previous exercise {@sec-anova}, you conducted an ANOVA

In this workshop, you will test whether two experimental factors — Light and Fertiliser — interact to affect plant biomass.
You’ll use a factorial dataset and compare two models:

- **Additive model** – assumes effects of Light and Fertiliser simply add together

- **Interaction model** – allows the effect of one variable to depend on the other

## Learning Outcomes

By completing this assignment, you will:

- Fit factorial ANOVA models in R using `lm()`.

- Test for an interaction term with `anova()`.

- Interpret model summaries and visualise predicted means.

- Decide when an interaction term is warranted.



### Set up your Quarto document

- Create a new .qmd file with a clear title, author, and date.

````md
---
title: "My Report"
author: "Your Name"
format: html
  embed-resources: true
execute:
  echo: false
  warning: false
  message: false

---

````

- Load packages

- Read in the dataset

- Complete the following sections


````md
```{r}`r ''`
#| include: false

library(tidyverse)
library(here)
library(rstatix)     # simple t-tests with tidy output
library(broom)       # tidy model summaries (optional)
library(sandwich)    # robust (heteroskedasticity-consistent) variance estimators
library(lmtest)      # coeftest/coefci for robust inference
library(performance) # easy model checks

biomass <- read_csv(here("data", "biomass.csv"))

## The data

## Visualisation

## Additive model

## Interaction model

## Model predictions

## Visualise model predictions

```
````

## The data

The dataset records biomass production (g/m²) for grassland plots treated with combinations of Fertiliser and Light.

Download and read the data

> How many observations are there, and what are the three variables in the dataset?


## Visualise treatment effects

Create a boxplot to explore how biomass varies with Fertiliser and Light.

> What pattern do you notice? Does the combined (F+ L+) treatment appear additive or possibly interactive?

## Fit an additive model

Now test whether each treatment affects biomass independently.

```{r, eval = FALSE}

model_add <- lm(Biomass.m2 ~ Fert + Light, data = biomass)
summary(model_add)

```

> Which treatments are significant predictors of biomass?

> What does the intercept represent in this model?

## Fit an interaction model

Add an interaction term and compare models.

```{r, eval = FALSE}

model_int <- lm(Biomass.m2 ~ Fert + Light +
                  Fert:Light, # interaction term
                data = biomass)
summary(model_int)

anova(model_add, model_int)

```

> Does the interaction term improve model fit (look at the F-test)?

> What does a significant interaction mean in this context?


## Visualise model predictions

- Use `emmeans`

- Plot predicted means and 95% confidence intervals


> Describe what the plot shows.

> Does Light increase biomass more when Fertiliser is present or absent?


## Check model assumptions

```{r, eval = FALSE}

check_model(model_int,
            detrend = FALSE)

```

> Do the residuals appear normally distributed and homoscedastic?

> Would you keep this model based on the diagnostics?

## Summarise your findings

In 4–6 sentences, interpret your analysis:

- Write a short results summary - it should include estimates of difference, uncertainty values, test statistics and p-values.

- Which model (additive or interaction) fits the data better?

- Summarise the biological conclusion in two sentences.


## Rendering your Quarto doc

After completing your analysis, render your `.qmd` file to Pdf to produce a polished report. Check the following:

- Code runs top to bottom – all chunks execute without errors.

- Figures and tables appear correctly – plots are visible, captions make sense.

- Formatting is clear – headers, text, and code chunks are readable.

- Interpretation is included – your narrative explains results, not just code output.
    
    - Include a figure
    - Include a short results section (with a model write-up and correlation test)

:::{.callout-warning collapse = "true"}
## My document won't render

If the document doesn’t render, check that all packages are loaded and all objects are created in previous chunks.

Check out {@sec-reports} for more help.
:::
